---
title: "Project Phase 1 Document"
format: html
editor: visual
---

## Data Analytics Group Project

**Description of Data set:**

**Cleanse Data:**

```{r}
initial_data <- read.csv("heart_failure_clinical_records_dataset.csv")

```

Check for missing values:

```{r}
mean(initial_data$age)
mean(initial_data$anaemia)
mean(initial_data$creatinine_phosphokinase)
mean(initial_data$diabetes)
mean(initial_data$ejection_fraction)
mean(initial_data$high_blood_pressure)
mean(initial_data$platelets)
mean(initial_data$serum_creatinine)
mean(initial_data$serum_sodium)
mean(initial_data$sex)
mean(initial_data$smoking)
mean(initial_data$time)
mean(initial_data$DEATH_EVENT)
```

This means that there are no missing values in the data set, because if there were missing values, then the mean would result in a value of NA for the given attribute.

Check for outliers:

```{r}
boxplot(initial_data$age)
boxplot(initial_data$creatinine_phosphokinase)
boxplot(initial_data$ejection_fraction)
boxplot(initial_data$platelets)
boxplot(initial_data$serum_creatinine)
boxplot(initial_data$serum_sodium)
boxplot(initial_data$time)
```

While there are several outliers in some of the attributes, I would not make sense to remove them from the data set. This is because there is a large quantity of outliers, and removing them would have a significant effect on the total amount of data in the data set.

**Analysis of Attributes:**

Age:

```{r}

```

Anemia:

Creatinine Phosphokinase:

Diabetes:

Ejection Fraction:

High Blood Pressure:

Platelets:

Serum Creatine:

Serum Sodium:

Sex:

Smoking:

Time:

Death Event:
